---
title: "Figures and Tables"
author: "Baijia Xu"
date: "04/14"
output:
  html_document:
    code_folding: hide
    self_contained: true
---

```{r}
library(dplyr)
library(ggplot2)
library(kableExtra)
```

## Simulation Study

### Local Demo

```{r source-functions}
source(here::here("R", "em_functions.R"))
source(here::here("R", "simulate_data.R"))
```

```{r simulate-demo}
# simulate data
dat = sim_lcr_data(n = 100, p = 6, K = 2, seed = 42)

cat(sprintf("n = %d, p = %d, q = %d, K = %d\n", nrow(dat$s_mat), dat$p, dat$q, dat$K))
cat(sprintf("True state frequencies: %s\n", paste(table(dat$true_z), collapse = ", ")))
```

```{r fit-em-demo}
# Best-of-20 fit (K-means init): used for all subsequent inference
fit = em_lcr(dat$s_mat, dat$r_mat, K = 2, n_init = 20, tol = 1e-4, max_iter = 500)

cat(sprintf("Converged: %s in %d iterations\n", fit$converged, fit$n_iter))
cat(sprintf("Final log-likelihood: %.2f\n", fit$loglik[fit$n_iter]))
```

### Log-Likelihood Convergence

To illustrate EM convergence dynamics, we run a single instance from a random initialization (not the best-of-20 K-means fit above). This shows the full path from a cold start to convergence.

```{r ll-convergence, fig.height=4, fig.width=7}
# Single run from random initialization — shows convergence path
fit_conv = em_lcr(dat$s_mat, dat$r_mat, K = 2, n_init = 1,
                  init_method = "random", tol = 1e-4, max_iter = 500, min_iter = 5)

tibble(
  iteration = 1:fit_conv$n_iter,
  loglik = fit_conv$loglik
) %>%
  ggplot(aes(x = iteration, y = loglik)) +
  geom_line(linewidth = 0.8, color = "#2166AC") +
  geom_point(size = 1.5, color = "#2166AC") +
  labs(
    title = "EM Log-Likelihood Convergence (single random initialization)",
    subtitle = sprintf("Converged in %d iterations", fit_conv$n_iter),
    x = "Iteration",
    y = "Observed Log-Likelihood"
  )
```

### Parameter Recovery

```{r param-recovery}
# align estimated parameters to true parameters
aligned = align_labels(fit$params, dat$true_params, K = 2)

# Each state has q = p(p+1)/2 = 21 beta0 and 21 beta1 values (one per edge).
# The diagonal simplification means edge j's task connectivity depends only on
# edge j's rest connectivity -- no cross-edge effects.
# We show the first 10 edges here; the scatter plot below shows all q edges.
n_show = min(10, dat$q)

param_table = bind_rows(
  tibble(
    state      = "State 1",
    edge       = 1:n_show,
    beta0_true = round(dat$true_params$beta0[1, 1:n_show], 4),
    beta0_est  = round(aligned$beta0[1, 1:n_show], 4),
    beta1_true = round(dat$true_params$beta1[1, 1:n_show], 4),
    beta1_est  = round(aligned$beta1[1, 1:n_show], 4)
  ),
  tibble(
    state      = "State 2",
    edge       = 1:n_show,
    beta0_true = round(dat$true_params$beta0[2, 1:n_show], 4),
    beta0_est  = round(aligned$beta0[2, 1:n_show], 4),
    beta1_true = round(dat$true_params$beta1[2, 1:n_show], 4),
    beta1_est  = round(aligned$beta1[2, 1:n_show], 4)
  )
)

param_table %>%
  kable(
    caption = sprintf(
      "Estimated vs. True Parameters: first %d of q=%d edges per state. Each state has q intercepts (beta0) and q slopes (beta1), one per edge.",
      n_show, dat$q),
    col.names = c("State", "Edge", "True beta0", "Est. beta0", "True beta1", "Est. beta1")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

```{r param-scatter, fig.height=5, fig.width=6}
# scatter plot of all estimated vs true beta1 values
plot_df = tibble(
  true = c(dat$true_params$beta1[1,], dat$true_params$beta1[2,]),
  estimated = c(aligned$beta1[1,], aligned$beta1[2,]),
  state = rep(c("State 1", "State 2"), each = dat$q)
)

ggplot(plot_df, aes(x = true, y = estimated, color = state)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    title = "Estimated vs. True Slope Parameters",
    x = expression("True " * beta[1]),
    y = expression("Estimated " * hat(beta)[1]),
    color = "Latent State"
  ) +
  coord_equal()
```

```{r mixing-props}
cat(sprintf("True mixing proportions: %s\n",
            paste(round(dat$true_params$pi, 3), collapse = ", ")))
cat(sprintf("Estimated mixing proportions: %s\n",
            paste(round(aligned$pi, 3), collapse = ", ")))
```

```{r clustering-accuracy}
# classification accuracy
gamma = em_estep(dat$s_mat, dat$r_mat, fit$params)
z_hat = apply(gamma, 1, which.max)

cat(sprintf("Estimated state frequencies: %s\n", paste(table(z_hat), collapse = ", ")))
```

### Model Selection

```{r model-selection}
sel = select_K(dat$s_mat, dat$r_mat, K_max = 4, n_init = 10)

sel$table %>%
  mutate(across(c(loglik, AIC, BIC), ~round(.x, 2))) %>%
  kable(caption = "Model Selection: AIC and BIC across K",
        col.names = c("K", "Log-lik", "# Params", "AIC", "BIC", "Converged", "Iterations")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

cat(sprintf("BIC-selected K: %d (true K = 2)\n",
            sel$table$K[which.min(sel$table$BIC)]))
```


### Full Simulation Study (Cluster)

The simulation is run via:

```bash
cd "final project"
sbatch simulations/run_sim.sh
```

## Results

### Bootstrap Inference

```{r run-bootstrap, cache=FALSE}
set.seed(2026)
B_boot = 200

boot_params = vector("list", B_boot)
n_boot = nrow(dat$s_mat)

for (b in 1:B_boot) {
  idx = sample(n_boot, n_boot, replace = TRUE)
  s_b = dat$s_mat[idx, , drop = FALSE]
  r_b = dat$r_mat[idx, , drop = FALSE]

  fit_b = em_lcr(s_b, r_b, K = 2, n_init = 1,
                 init_params = fit$params, tol = 1e-5, max_iter = 300)

  if (fit_b$converged || fit_b$n_iter > 10) {
    aligned_b = align_labels(fit_b$params, fit$params, K = 2)
    boot_params[[b]] = aligned_b
  }
}

# drop any NULL (non-converged)
boot_params = Filter(Negate(is.null), boot_params)
B_actual = length(boot_params)
cat(sprintf("Successful bootstrap replicates: %d / %d\n", B_actual, B_boot))
```

```{r bootstrap-pi-table}
# mixing proportion inference
pi_full = fit$params$pi
pi_boot_mat = do.call(rbind, lapply(boot_params, `[[`, "pi"))

pi_ci = apply(pi_boot_mat, 2, quantile, probs = c(0.025, 0.975))
pi_se = apply(pi_boot_mat, 2, sd)

pi_summary_demo = data.frame(
  State     = paste("State", 1:2),
  Estimate  = round(pi_full, 4),
  SE        = round(pi_se, 4),
  CI_Lower  = round(pi_ci[1, ], 4),
  CI_Upper  = round(pi_ci[2, ], 4)
)

pi_summary_demo %>%
  kable(caption = sprintf("Bootstrap Inference: Mixing Proportions (%d replicates, n=100, p=6, K=2)", B_actual),
        col.names = c("State", "Estimate", "Bootstrap SE", "95% CI Lower", "95% CI Upper")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r bootstrap-beta1-table}
# beta1 inference for first 8 edges per state
beta1_full = fit$params$beta1
beta1_boot_arr = array(
  do.call(rbind, lapply(boot_params, function(x) c(x$beta1))),
  dim = c(B_actual, 2, dat$q)
)

n_show_b = 8
beta1_rows = list()
for (k in 1:2) {
  for (j in 1:n_show_b) {
    bv = beta1_boot_arr[, k, j]
    beta1_rows[[length(beta1_rows) + 1]] = data.frame(
      State    = paste("State", k),
      Edge     = j,
      Estimate = round(beta1_full[k, j], 4),
      SE       = round(sd(bv), 4),
      CI_Lower = round(quantile(bv, 0.025), 4),
      CI_Upper = round(quantile(bv, 0.975), 4)
    )
  }
}
beta1_table = do.call(rbind, beta1_rows)

beta1_table %>%
  kable(caption = "Bootstrap Inference: Slope Parameters beta1 (first 8 edges per state)",
        col.names = c("State", "Edge", "Estimate", "Bootstrap SE", "95% CI Lower", "95% CI Upper")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

```{r bootstrap-dist-plot, fig.height=4, fig.width=8}
# Bootstrap distribution of pi_1 and beta1[1,1]
pi1_boot = pi_boot_mat[, 1]
b1_edge1_boot = beta1_boot_arr[, 1, 1]

plot_df_boot = bind_rows(
  tibble(param = "pi_1 (mixing proportion, State 1)", value = pi1_boot,
         estimate = pi_full[1]),
  tibble(param = "beta1_1 (slope, State 1, Edge 1)", value = b1_edge1_boot,
         estimate = beta1_full[1, 1])
)

ggplot(plot_df_boot, aes(x = value)) +
  geom_histogram(bins = 40, fill = "#4393C3", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = estimate), color = "#D6604D", linewidth = 1, linetype = "dashed") +
  facet_wrap(~param, scales = "free") +
  labs(
    title = sprintf("Bootstrap Distributions (%d replicates)", B_actual),
    subtitle = "Red dashed line = full-data estimate",
    x = "Bootstrap estimate", y = "Count"
  )
```


### Simulation Study Results

```{r load-sim-results}
sim_files = list.files(here::here("results", "20260414"), pattern = "\\.RDA$",
                       full.names = TRUE)

all_results = list()
for (f in sim_files) {
  load(f)
  all_results = c(all_results, results)
}

sim_df = bind_rows(all_results) %>%
  mutate(
    n = factor(n, levels = c(50, 100, 200)),
    p_label = paste0("p = ", p, " (q = ", p*(p+1)/2, ")"),
    K_label = paste0("K = ", K_true)
  )
```

```{r sim-summary-table}
sim_df %>%
  group_by(n, p, K_true) %>%
  summarise(
    bic_accuracy   = mean(bic_correct),
    mean_beta1_err = mean(beta1_frob_err),
    mean_ari       = mean(ari),
    pct_converged  = mean(converged) * 100,
    .groups = "drop"
  ) %>%
  arrange(K_true, p, n) %>%
  mutate(across(where(is.numeric), ~round(.x, 3))) %>%
  kable(
    caption = "Simulation Study Summary (500 replicates per scenario)",
    col.names = c("n", "p", "True K", "BIC Accuracy",
                  "Mean ‖beta1 error‖_F", "Mean ARI", "% Converged")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  pack_rows("K = 2", 1, 6) %>%
  pack_rows("K = 3", 7, 12) %>%
  pack_rows("K = 4", 13, 18)
```

```{r sim-bic-plot, fig.height=5, fig.width=9}
sim_df %>%
  group_by(n, p_label, K_label) %>%
  summarise(bic_accuracy = mean(bic_correct), .groups = "drop") %>%
  ggplot(aes(x = n, y = bic_accuracy, color = p_label, group = p_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~K_label) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_color_manual(values = c("#2166AC", "#D6604D")) +
  labs(
    title = "BIC Model Selection Accuracy by Sample Size, Dimensionality, and True K",
    x = "Sample size (n)", y = "BIC accuracy (proportion correct)",
    color = "Dimensionality"
  ) +
  theme(legend.position = "bottom")
```

```{r sim-ari-plot, fig.height=5, fig.width=9}
sim_df %>%
  ggplot(aes(x = n, y = ari, fill = p_label)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.7, position = position_dodge(width = 0.8)) +
  facet_wrap(~K_label) +
  scale_fill_manual(values = c("#4393C3", "#F4A582")) +
  labs(
    title = "Clustering Accuracy (Adjusted Rand Index) by Scenario",
    x = "Sample size (n)", y = "ARI",
    fill = "Dimensionality"
  ) +
  theme(legend.position = "bottom")
```

```{r sim-error-plot, fig.height=5, fig.width=9}
sim_df %>%
  group_by(n, p_label, K_label) %>%
  summarise(mean_err = mean(beta1_frob_err),
            se_err   = sd(beta1_frob_err) / sqrt(n()),
            .groups  = "drop") %>%
  ggplot(aes(x = n, y = mean_err, color = p_label, group = p_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_err - 1.96 * se_err,
                    ymax = mean_err + 1.96 * se_err), width = 0.1) +
  facet_wrap(~K_label, scales = "free_y") +
  scale_color_manual(values = c("#2166AC", "#D6604D")) +
  labs(
    title = "Parameter Estimation Error: Mean Frobenius Norm of beta1",
    subtitle = "Error bars = +/- 1.96 x Monte Carlo SE",
    x = "Sample size (n)", y = "Mean Frobenius error",
    color = "Dimensionality"
  ) +
  theme(legend.position = "bottom")
```
